Improving Emotion Detection through Real-Time Translation of Text to ML Models Trained in Different Languages

### 2023-07-18 paper/appendix/code.tex
Paper Updates

\section{Appendix -- Application \& Code Details}
\subsection{Application}
Hello World - Application \& Code Details

\subsection{Code (Python)}

>>>>

\section{Appendix -- Application \& Code Details}
Hello World - Application \& Code Details
\subsection{Application}
\label{appendix:code_application}
Hello World - Application
\clearpage

\subsection{Code Details}
Hello World - Code Details

\subsection{RESTful API Server}
\label{appendix:code-api-server}
The RESTful\footnote{RESTful API is an interface that two computer systems use to exchange information securely over the internet.} API Server will be a simple Python Flask server listening on port \texttt{8080} on the local machine.

### 2023-07-18 paper/appendix/dataset.tex
Paper Updates

\subsection{Using PySpark to Load Data}
All example code was created using Jupyter Notebooks creating a PySpark session. The deails of setting up such session are below in the code example. All subsequent details are derived from the \texttt{sc}\footnote{sc = Spark Session}.
\begin{minted}{python}
# Setup Spark Session
# Spark Session Name: Qualyfing_Exam
from pyspark.sql import SparkSession

# Get Imports Needed
from pyspark.sql.functions import col, udf

@@

# Setup Spark Session
sc = SparkSession \
        .builder \
        .master("local[*]") \
        .appName("Qualyfing_Exam") \
        .getOrCreate()
\end{minted}

\subsection{Details on English Dataset}
The English dataset based on tweets\footnote{Tweets from Twitter} are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file. The following are some simply details on the laod and processing of such file.
\begin{minted}{python}
# Read csv file to Saprk dataframe
df_english = sc.read.csv("data/english.csv", header=True, inferSchema=True)
# Group by a column and count the occurrences in df_english
count_english = df_english.groupBy('sentiment').count()

# Show the result for df_english
count_english.show()
+----------+-----+
| sentiment|count|
+----------+-----+
|   boredom|  179|
|      love| 3842|
|    relief| 1526|
|       fun| 1776|
|      hate| 1323|
|   neutral| 8638|
|     anger|  110|
| happiness| 5209|
|  surprise| 2187|
|   sadness| 5165|
|     worry| 8459|
|enthusiasm|  759|
|     empty|  827|
+----------+-----+

\end{minted}

\subsection{Details on German Dataset}
The German dataset was downloaded from the ETH's\footnote{ETH (German: Eidgenössische Technische Hochschule) or Swiss Federal Institute of Technology} servers via "link" as a \texttt{JSON} file.
\begin{minted}{python}
# Read JSON file to Saprk dataframe
df_german = sc.read.json("data/german.json")

# Group by a column and count the occurrences in df_german
count_german = df_german.groupBy('article_emotion').count()

# Show the result for df_german
count_german.show(truncate=0)
+----------------------------------+-----+
|article_emotion                   |count|
+----------------------------------+-----+
|[Ärger, Antizipation, Traurigkeit]|1    |
|[Überraschung, Vertrauen]         |14   |
|[Überraschung, Freude]            |15   |
|[Ärger, Vertrauen]                |7    |
|[Ekel, Traurigkeit]               |4    |
|[Keine, Antizipation]             |5    |
|[Vertrauen]                       |130  |
|[Freude]                          |58   |
|[Ärger, Antizipation]             |31   |
|[Ärger]                           |114  |
|[Überraschung]                    |154  |
|[Antizipation, Traurigkeit]       |15   |
|[Antizipation, Freude, Vertrauen] |2    |
|[Vertrauen, Angst]                |5    |
|[Traurigkeit]                     |101  |
|[Antizipation]                    |413  |
|[Traurigkeit, Vertrauen]          |7    |
|[Antizipation, Vertrauen]         |104  |
|[Ärger, Traurigkeit, Angst]       |1    |
|[Unklar]                          |314  |
+----------------------------------+-----+

\end{minted}

>>>>

\subsection{Using PySpark to Load Data}
All example code was created using Jupyter Notebooks creating a PySpark session. The details of setting up such session are below in the code example. All subsequent details are derived from the \texttt{sc}\footnote{sc = Spark Session}.
\begin{minted}[fontsize=\small]{python}
# General Imports
import glob
import shutil

# Setup Spark Session
from pyspark.sql import SparkSession

# Get Spark Functions Needed
from pyspark.sql.functions import col, udf, split, explode

@@

# Setup Spark Session
sc = SparkSession \
        .builder \
        .master("local[*]") \
        .appName("data_clean_up") \
        .getOrCreate()

# Print Spark Version being run
print("Spark V: ", sc.version)
\end{minted}
\clearpage

\subsection{Details on English Dataset}
\label{appendix:dataset_english}
The English dataset is based on tweets\footnote{Tweets from Twitter} that are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file. The following code was use to filter and process the file for use in this project.
\begin{minted}[fontsize=\small]{python}
# *********************************
# *** English Data Preparations ***
# *********************************

# Load English CSV File into a Dataframe
df_english = sc.read.csv("data/english.csv", header=True, inferSchema=True)
print(f'Count (raw): {df_english.count()}')

# Print Columns
print(f'English Source Columns: {df_english.columns}\n')

# Filter only lavbels that we mathc with the German counter parts
filter_values = ["love", "hate", "neutral", "anger", "happiness", "surprise", "sadness", "worry", "enthusiasm"]
df_english_filtered = df_english.filter(col("sentiment").isin(filter_values))

# Remove unnecessary column
df_english_filtered = df_english_filtered.drop("tweet_id")

# Rename Columns
df_english_filtered = df_english_filtered.withColumnRenamed("sentiment", "emotion")
df_english_filtered = df_english_filtered.withColumnRenamed("content", "sentence")

# Make sure the column order to the same for both german and english csv files
df_english_filtered = df_english_filtered.select('sentence', 'emotion')

print(f'Count (filtered): {df_english_filtered.count()}')

# Group By for Details & Count
df_english_grouped = df_english_filtered.groupBy('emotion').count()

# Show Groupings and Respetive Counts
print('\nGrouped & Count by "emotion"')
df_english_grouped.show(truncate=0)

# Save Dataframe to CSV
directory_path = 'data/spark_data_parts'
df_english_filtered.coalesce(1).write.csv(directory_path, header=True, mode="overwrite")

file_pattern = 'part-00000*.csv'
file_path = glob.glob(directory_path + '/' + file_pattern)[0]

shutil.move(file_path, './data/data_en.csv')
\end{minted}
\clearpage

\subsection{Details on German Dataset}
\label{appendix:dataset_german}
The German dataset was downloaded from the ETH's\footnote{ETH (German: Eidgenössische Technische Hochschule) or Swiss Federal Institute of Technology} servers via "link" as a \texttt{JSON} file. Since the the file was \texttt{JSON} we will need to filter and explode the file based on teh emotions that are attached to it.
\begin{minted}[fontsize=\small]{python}
# ********************************
# *** German Data Preparations ***
# ********************************

# Load German JSON File into a Dataframe
df_german = sc.read.json("data/german.json")
print(f'Count (raw): {df_german.count()}')

# Print Columns
print(f'German Source Columns: {df_german.columns}\n')

# We Need to "split" the "artice_emotion" column since the ETH team listed
# multiple emotions in one column
# In order to "explode" the column to it's distinct rows
df_german_exploded = df_german.select('*', explode('article_emotion').alias('emotion'))

# Rename Column
df_german_exploded = df_german_exploded.withColumnRenamed("title", "sentence")

# Remove unnecessary column
df_german_exploded = df_german_exploded.drop("article_id")
df_german_exploded = df_german_exploded.drop("article_stance")
df_german_exploded = df_german_exploded.drop("paragraphs")
df_german_exploded = df_german_exploded.drop("source")
df_german_exploded = df_german_exploded.drop("article_emotion")
df_german_exploded = df_german_exploded.drop("snippet")

# Make sure the column order to the same for both german and english csv files
df_german_exploded = df_german_exploded.select('sentence', 'emotion')

print(f'Count (filtered): {df_german_exploded.count()}')

# Group By for Details & Count
df_german_grouped = df_german_exploded.groupBy('emotion').count()

# Show Groupings and Respetive Counts
print('\nGrouped & Count by "emotion"')
df_german_grouped.show(truncate=0)

# Save Dataframe to CSV
directory_path = 'data/spark_data_parts'
df_german_exploded.coalesce(1).write.csv(directory_path, header=True, mode="overwrite")

file_pattern = 'part-00000*.csv'
file_path = glob.glob(directory_path + '/' + file_pattern)[0]

shutil.move(file_path, './data/data_de.csv')
\end{minted}

### 2023-07-18 paper/main.tex
Paper Updates

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{fullpage}
\usepackage{multirow}

\usepackage{minted}
\setminted[]{linenos, samepage, numbersep=5pt, frame=lines}

@@

\begin{wrapfigure}{r}{0.5\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.5\textwidth]{Plutchik-Wheel}
    \caption{Plutchik Wheel of Emotions}
    \label{fig:plutchik-wheel}
\end{wrapfigure}

@@

\begin{itemize}
\item Joy: A feeling of happiness, contentment, or delight.
\item Sadness: A state of unhappiness, sorrow, or grief.
\item Anger: An intense emotional response associated with feelings of displeasure, frustration, or hostility.
\item Fear: An emotional response triggered by perceived threats or danger, often associated with anxiety or terror.
\item Surprise: A sudden and unexpected reaction to something unusual or startling.
\item Disgust: A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
\item Trust: A sense of reliance, confidence, or faith in someone or something.
\item Anticipation: A feeling of excited expectation or eagerness towards future events.
\end{itemize}
It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions. However for the scope of this research project the most common 7-8 emotions are used as described above.

\subsection{Background and significance of Emotion Detection (ED) analysis in Machine Learning models}
Emotion Detection (ED) analysis in Machine Learning models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific target\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to name a few.

By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.

@@

Due to the before mentioned this field of research provides significance in enabling machines to comprehend and respond appropriately to human emotions, contributing to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications that are growing rapidly.

\subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes is much smaller. The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans. To compound on the issue of the lack of data; the many datasets that are available are in multiple languages.

\subsection{Motivation for the research project focusing on extending ED datasets through translation and evaluating the impact on prediction rates}
With the before mentioned issues regarding the quantity/lack and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The hope is that by use of translation that the predictability of ED can be improved by use of real- real-time language conversion and processing.

@@

\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.4\textwidth]{Dataset-Extension}
    \caption{Example of English dataset extension by translating from German.}
    \label{fig:dataset-extension}
\end{wrapfigure}

The procurement of two ED datasets is needed. The first will be in English and the second in German. Each dataset will be translated in the other’s language for extending purposes. As an example figure \ref{fig:dataset-extension}, the English will be translated to German and the German to English by which the original datasets are extended.

Once the two datasets have been translated, we now extend the original language dataset with the translated once. In essence extending the datasets to create a larger one. With both extended datasets we can now train our ML models for English and German.

\subsubsection{PySpark Application Development for Translation \& Prediction}
A program will be built in Jupiter notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English. Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and compared. The average of the prediction will be used for final prediction of the emotion that was processed.

@@

    \hline
    \multicolumn{2}{|c|}{English} & \multicolumn{2}{|c|}{German} & \multirow{2}{*}{Used} \\
    Name & Count & Name & Count \\
    \hline
    Boredom    &  179 & ---           & --- & NO  \\
    Love       & 3842 & Vertrauen     & 130 & YES \\
    Relief     & 1526 & ---           & --- & NO  \\ 
    Fun        & 1776 & ---           & --- & NO  \\
    Hate       & 1323 & Ekel          &   4 & YES \\
    Neutral    & 8638 & Unklar        & 314 & YES \\
    Anger      &  110 & Ärger         & 114 & YES \\
    Happiness  & 5209 & Freude        &  58 & YES \\
    Surprise   & 2187 & Überraschung  & 154 & YES \\
    Sadness    & 5165 & Traurigkeit   & 101 & YES \\
    Worry      & 8459 & Angst         &   6 & YES \\
    Enthusiasm &  759 & Antizipation  & 413 & YES \\
    Empty      &  827 & ---           & --- & NO  \\
    \hline
\end{tabular}
\caption{Dataset of English \& German and their respective label counts.}
\label{table:dataset_labels}
\end{table}

@@

\begin{quote}
\begin{minted}{python}
# Using Deep Translator to leveage Google Translate
# Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
from deep_translator import GoogleTranslator

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[toc,page]{appendix}
\usepackage{microtype}
\usepackage{fullpage}
\usepackage{multirow}

\usepackage{minted}
\setminted[]{linenos, numbersep=5pt, frame=lines}

@@

\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.35\textwidth]{Plutchik-Wheel}
    \caption{Plutchik Wheel of Emotions}
    \label{fig:plutchik-wheel}
\end{wrapfigure}

@@

\begin{itemize}
\item Joy (Fun): A feeling of happiness, contentment, or delight.
\item Sadness: A state of unhappiness, sorrow, or grief.
\item Anger: An intense emotional response associated with feelings of displeasure, frustration, or hostility.
\item Worry: An emotional response triggered by perceived threats or danger, often associated with anxiety or terror.
\item Surprise: A sudden and unexpected reaction to something unusual or startling.
\item Disgust (Hate): A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
\item Trust: A sense of reliance, confidence, or faith in someone or something.
\item Anticipation: A feeling of excited expectation or eagerness towards future events.
\end{itemize}
It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions. However for the scope of this research project the most common 8 emotions are used as described above. Both the English and German datasets being used in this research matched to the 8 listed above. Details on the linkage of the two datasets is described in the next section with an emphasis on the Table \ref{table:dataset_labels} detailed linkage.

\subsection{Background and significance of Emotion Detection (ED) analysis in Machine Learning models}
Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to just name a few.

By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}. With the goal in mind of improving ML models this research paper came to be.

@@

Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human emotions. This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on.

\subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller. The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans. To compound on the issue of the lack of data; the many datasets that are available are in multiple languages.

\subsection{Motivation for the research project focusing on extending ED datasets through translation and evaluating the impact on prediction rates}

This paper is significant because an entire industry of automated purported emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. The market for emotion recognition software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, and a multiple of others, it is fitting to work on research to solve issues.

With the before mentioned issues regarding the quantity/lack and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The hope is that by use of translation that the predictability of ED can be improved by use of real- real-time language conversion and processing.

@@

\begin{wrapfigure}{R}{0.4\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
    \caption{Example of English dataset extension by translating from German.}
    \label{fig:dataset-extension}
\end{wrapfigure}

The procurement of two ED datasets is needed. The first will be in English and the second in German. Each dataset will be translated into the other’s language for extending purposes. As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.

The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.

Once the two datasets have been translated, we now extend the original language dataset with the translated one.By this approach we are in essence extending the original datasets to create a larger one wiht the foreign language data. With both extended datasets we can now train our ML models for English and German.

\subsubsection{PySpark Application Development for Translation \& Prediction}
A program will be built in Jupiter Notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English. Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and will be compared to each other's output. The average of the prediction will be used for final prediction of the supplied sentence and the emotion that was detected.

In order for such to be processed in real-time, the Jupyter Notebook will start a PySpark session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.

@@

    \hline
    \multicolumn{2}{|c|}{English} & \multicolumn{2}{|c|}{German} & \multirow{2}{*}{Used} \\
    Name & Count & Name & Count \\
    \hline
    Boredom    &  179 & ---           & --- & NO  \\
    Love       & 3842 & Vertrauen     & 316 & YES \\
    Relief     & 1526 & ---           & --- & NO  \\ 
    Fun        & 1776 & ---           & --- & NO  \\
    Hate       & 1323 & Ekel          &  29 & YES \\
    Neutral    & 8638 & Unklar        & 314 & YES \\
    Anger      &  110 & Ärger         & 226 & YES \\
    Happiness  & 5209 & Freude        & 140 & YES \\
    Surprise   & 2187 & Überraschung  & 369 & YES \\
    Sadness    & 5165 & Traurigkeit   & 184 & YES \\
    Worry      & 8459 & Angst         & 154 & YES \\
    Enthusiasm &  759 & Antizipation  & 774 & YES \\
    Empty      &  827 & ---           & --- & NO  \\
    \hline
\end{tabular}
\caption{Dataset of English \& German and their respective label counts.}
\label{table:dataset_labels}
\end{table}

Details on how we formatted the the two (English and German) datasets to each other can be viewed in the Appendix \ref{appendix:dataset_english} and \ref{appendix:dataset_german}

@@

\begin{quote}
\begin{minted}[fontsize=\small]{python}
# Using Deep Translator to leveage Google Translate
# Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
from deep_translator import GoogleTranslator
